---
title: "Analysis Code"
format: html
---

## Setup

```{r}
## open packages:
library(tidyverse)
library(invburreg)
library(coolorrr)
library(poweRlaw)
library(geomtextpath)
library(ggregtab)

## session options:
data("wars")
set_palette(
  qualitative = c("steelblue", "gray", "red3", "navy", "orange3")
)
theme_set(ggthemes::theme_fivethirtyeight())
theme_update(
  panel.background = element_rect(fill = "white"),
  plot.background = element_rect(fill = "white"), 
  legend.key = element_rect(fill = "white"),
  plot.caption = element_text(
    hjust = 0,
    size = 12
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 12
  ),
  plot.caption.position = "plot",
  axis.title = element_text(
    size = 12,
    hjust = .5
  )
)
saveit <- function(num = 1, ht = 4, wd = 6) {
  ggsave(
    filename = here::here(
      "_figs",
      paste0("fig", num, ".png")
    ),
    dpi = 500,
    height = ht,
    width = wd
  )
}
```

## Figure 1

Figure 1 shows variation in war sizes over time.

```{r}
#| fig-height: 4
#| fig-width: 6
ggplot(wars) +
  aes(x = year, y = fat) +
  geom_point() +
  labs(
    x = "Start Year",
    y = "Total Battle Deaths",
    caption = str_wrap(
      "Figure 1: The total number of battle deaths by war in the Correlates of War conflict series. Values shown by war start year.",
      width = 75
    )
  ) +
  scale_x_continuous(
    breaks = seq(1820, 2010, by = 20)
  ) +
  scale_y_continuous(
    labels = ~ paste0(.x / 1e06, " mil")
  )
saveit(1)
```

## Figure 2

To fit the power-law to data I'm using the `{poweRlaw}` R package.

```{r}
## fit the power-law to conflict deaths
pl <- conpl$new(wars$fat)
pl$setXmin(
  estimate_xmin(
    pl, 
    xmax = max(wars$fat)
  )
)
```

Visualize the model:

```{r}
#| fig-height: 4
#| fig-width: 6
p <- function(x) {
  rank(-x) / max(rank(-x))
}
wars |> 
  mutate(
    pfat = p(fat)
  ) |>
  ggplot() +
  aes(fat, pfat) +
  geom_point(
    color = "gray"
  ) +
  geom_smooth(
    data = . %>% 
      filter(fat >= pl$xmin),
    method = "lm",
    se = F
  ) +
  geom_textvline(
    xintercept = pl$xmin,
    label = "Optimal data\nthreshold"
  ) +
  scale_x_log10(
    labels = ~ paste0(
      .x / 1e06, " mil"
    )
  ) +
  scale_y_log10(
    labels = scales::percent
  ) +
  labs(
    subtitle = expression(
      alpha*" = "*1.525*"; x-min = 7,061; data loss = 46.3%"
    ),
    x = "Battle Deaths (in millions)",
    y = "Pr(X > x)",
    caption = str_wrap(
      "Figure 2: The optimal power-law fit for the battle death series. Values are shown on the log-scale.",
      width = 75
    )
  ) 
saveit(2)
```


## Figure 3

The inverse burr fit for the data.

```{r}
base_fit <- ibm(
  outcome = fat,
  data = wars,
  its = 500
)
base_fit$out |>
  transmute(param, estimate = exp(estimate))
```

Visualize its fit for the data.

```{r}
#| fig-height: 4 
#| fig-width: 6

wars |> 
  mutate(
    pfat = p(fat)
  ) |>
  ggplot() +
  aes(fat, pfat) +
  geom_point(
    color = "gray"
  ) +
  geom_line(
    aes(
      y = actuar::pinvburr(
        q = fat,
        scale = exp(base_fit$out$estimate[1]),
        shape1 = exp(base_fit$out$estimate[2]),
        shape2 = exp(base_fit$out$estimate[3]),
        lower.tail = F
      )
    ),
    color = "blue",
    linewidth = 1
  ) +
  scale_x_log10(
    labels = ~ paste0(
      .x / 1e06, " mil"
    )
  ) +
  scale_y_log10(
    labels = scales::percent
  ) +
  labs(
    subtitle = expression(
      mu*" = 0.49; "*alpha*" = 438.29; "*theta*" = 0.67"
    ),
    x = "Battle Deaths (in millions)",
    y = "Pr(X > x)",
    caption = str_wrap(
      "Figure 3: The optimal inberse Burr fit for the battle death series. Values are shown on the log-scale.",
      width = 75
    )
  ) 
saveit(3)
```



## Figure 4

```{r}
#| fig-height: 4
#| fig-width: 6

px <- function(q = c(0, 10), 
               mu = 2, alpha = 2, theta = 2) {
  tibble(
    x = seq(q[1], q[2], len = 100000),
    y = actuar::dinvburr(
      x = x,
      scale = mu,
      shape1 = alpha,
      shape2 = theta
    )
  )
}

bind_rows(
  px(mu = .5) |> mutate(val = "0.5"),
  px(mu = 1) |> mutate(val = "1"),
  px(mu = 2) |> mutate(val = "2"),
  px(mu = 4) |> mutate(val = "4")
) |>
  ggplot() +
  aes(x, y, color = val) +
  geom_line(linewidth = 1) +
  labs(
    subtitle = expression(alpha*" and "*theta*" held constant at 2"),
    caption = str_wrap(
      "Figure 4: The influence of a change to the scale parameter on the inverse Burr distribution.",
      width = 75
    ),
    x = "x",
    y = "Denisity",
    color = expression(mu*" = ")
  ) +
  ggpal("diverging", ordinal = T, levels = 4)
saveit(4)
```

## Figure 5


```{r}
#| fig-height: 4
#| fig-width: 6

bind_rows(
  px(alpha = .5) |> mutate(val = "0.5"),
  px(alpha = 1) |> mutate(val = "1"),
  px(alpha = 2) |> mutate(val = "2"),
  px(alpha = 4) |> mutate(val = "4")
) |>
  ggplot() +
  aes(x, y, color = val) +
  geom_line(linewidth = 1) +
  labs(
    subtitle = expression(mu*" and "*theta*" held constant at 2"),
    caption = str_wrap(
      "Figure 5: The influence of a change to the first shape parameter on the inverse Burr distribution.",
      width = 75
    ),
    x = "x",
    y = "Denisity",
    color = expression(alpha*" = ")
  ) +
  ggpal("diverging", ordinal = T, levels = 4)
saveit(5)
```


## Figure 6

```{r}
#| fig-height: 4
#| fig-width: 6

bind_rows(
  px(theta = .5) |> mutate(val = "0.5"),
  px(theta = 1) |> mutate(val = "1"),
  px(theta = 2) |> mutate(val = "2"),
  px(theta = 4) |> mutate(val = "4")
) |>
  ggplot() +
  aes(x, y, color = val) +
  geom_line(linewidth = 1) +
  labs(
    subtitle = expression(alpha*" and "*mu*" held constant at 2"),
    caption = str_wrap(
      "Figure 6: The influence of a change to the second shape parameter on the inverse Burr distribution.",
      width = 75
    ),
    x = "x",
    y = "Denisity",
    color = expression(theta*" = ")
  ) +
  ggpal("diverging", ordinal = T, levels = 4)
saveit(6)
```



## Table 1

With covariates.

```{r}
preds <- ~ post1950 + dem + pop
cov_fit <- ibm(
  fat,
  mu = preds,
  alpha = preds,
  theta = preds,
  data = wars,
  its = 500
)
cov_fit$out |>
  select(param, term, estimate, statistic)
```


```{r}
#| fig-height: 4
#| fig-width: 6
bind_rows(
  base_fit$out |> mutate(model = "Baseline"),
  cov_fit$out |> mutate(model = "Covariates")
) |>
  mutate(
    conf.low = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    N = 95
  ) |>
  ggregtab(ratio = .75) +
  facet_wrap(
    ~ paste0("log(",param,")"),
    labeller = label_parsed
  ) +
  scale_y_discrete(
    labels = c("Population (log)", "Polity (Avg.)", "Post-1950",
               "Constant")
  ) +
  labs(
    subtitle = str_wrap(
      "Table 1: Inverse Burr regression estimates with bootstrapped standard errors.",
      width = 80
    )
  )
saveit("_regtab")
```


## Figure 7

A function to take the output from an inverse burr model and simulate random draws from the distribution.

```{r}
ibm_sim <- function(model, its = 1000, newdata = NULL) {
  if(is.null(newdata)) {
    pdata <- model$model_data
  } else {
    newdata <- as.data.frame(newdata)
    dnames <- colnames(model$model_data)
    ndata <- matrix(1, nrow(newdata), length(dnames))
    colnames(ndata) <- dnames
    for(i in 1:length(dnames)) {
      if(dnames[i] %in% colnames(newdata)) {
        ndata[, i] <- 
          newdata[, colnames(newdata) == dnames[i]]
      }
    }
    pdata <- ndata
  }
  param_fit <- function(model, data = pdata, par = "mu") {
    npars <- 1:nrow(model$out)
    names(npars) <- model$out$param
    exp(data[
      , npars[names(npars) == par]
    ] %*% (model$out |>
      filter(param == par) |>
      pull(estimate)))
  }
  map_dfr(
    .x = 1:(its * nrow(pdata)),
    .f = ~ newdata |>
      mutate(
        sim = .x,
        pred = actuar::rinvburr(
          n = nrow(pdata),
          scale = param_fit(model, par = "mu"),
          shape1 = param_fit(model, par = "alpha"),
          shape2 = param_fit(model, par = "theta")
        )
      )
  )
}
```


A function to get the cdf of an inverse burr model given fitted model parameters.

```{r}
ibm_d <- function(model, range = c(0, 1000), newdata = NULL) {
  if(is.null(newdata)) {
    pdata <- model$model_data
  } else {
    newdata <- as.data.frame(newdata)
    dnames <- colnames(model$model_data)
    ndata <- matrix(1, nrow(newdata), length(dnames))
    colnames(ndata) <- dnames
    for(i in 1:length(dnames)) {
      if(dnames[i] %in% colnames(newdata)) {
        ndata[, i] <- 
          newdata[, colnames(newdata) == dnames[i]]
      }
    }
    pdata <- ndata
  }
  param_fit <- function(model, data = pdata, par = "mu") {
    npars <- 1:nrow(model$out)
    names(npars) <- model$out$param
    exp(data[
      , npars[names(npars) == par]
    ] %*% (model$out |>
      filter(param == par) |>
      pull(estimate)))
  }
  x <- seq(range[1], range[2], len = 10000)
  p <- actuar::pinvburr(
          q = x,
          scale = param_fit(model, par = "mu"),
          shape1 = param_fit(model, par = "alpha"),
          shape2 = param_fit(model, par = "theta"),
          lower.tail = F
        )
  tibble(x, p)
}
```


```{r}
set.seed(111)
ibm_sim(
  base_fit,
  newdata = tibble(cont = 1),
) -> base_sim
ibm_sim(
  cov_fit,
  newdata = wars |> filter(warnum == 139) |>
    select(pop, dem, post1950)
) -> cov_sim


bind_rows(
  base_sim |>
    mutate(model = "Baseline"),
  cov_sim |>
    mutate(model = "World War II")
) |>
  group_by(model) |>
  mutate(
    ppred = p(pred)
  ) |>
  ggplot() +
  aes(pred, ppred, color = model) +
  geom_point() +
  geom_point(
    data = wars |> 
      mutate(pfat = p(fat)) |>
      filter(fat == max(fat)),
    aes(
      x = fat,
      y = pfat,
      color = "World War II (Actual)"
    ),
    size = 3
  ) +
  scale_x_log10(
    labels = ~ paste0(
      .x / 1e06, " mil"
    )
  ) +
  scale_y_log10(
    labels = scales::percent
  ) +
  labs(
    x = "Battle Deaths (in millions)",
    y = "Pr(X > x)",
    caption = str_wrap(
      "Figure 7: Simulated inverse Burr distributions based on the basline model and the model fit with covariates. In the covariate model, variables are set to World War II values. The actual Pr(X > x) for World War II is highlighted with a red point. Values are on the log-log scale",
      width = 75
    ),
    color = NULL
  ) +
  ggpal()
saveit(7)
```


## Figure 8

Show the conditional effect of democracy

```{r}
#| fig-height: 4
#| fig-width: 6

ibm_sim(
  cov_fit,
  newdata = tibble(
    post1950 = mean(wars$post1950),
    dem = c(-10, 0, 10),
    pop = mean(wars$pop)
  )
) -> dem_sim

dem_sim |>
  group_by(dem) |>
  mutate(ppred = p(pred)) |>
  ggplot() +
  aes(pred, ppred, color = as.factor(dem)) |>
  geom_point(alpha = 0.4) +
  scale_x_log10(
    labels = ~ paste0(.x / 1e06, " mil")
  ) +
  scale_y_log10(
    labels = scales::percent
  ) +
  ggpal() +
  labs(
    caption = str_wrap(
      "Figure 8: The conditional impact of average polity scores among belligerents on the CDF of war fatalities.",
      width = 75
    ),
    x = "Battle Deaths (in millions)",
    y = "Pr(X > x)",
    color = "Mean Polity"
  )
saveit(8)
```

## Table 2

```{r}
cov_fit0 <- ibm(
  fat,
  mu = preds,
  alpha = preds,
  theta = preds,
  data = wars |> mutate(fat = fat - 999) |>
    mutate(fat = ifelse(fat == 1, 1:sum(fat == 1), fat)),
  its = 500
)
```


```{r}
#| fig-height: 4
#| fig-width: 6
bind_rows(
  cov_fit$out |> mutate(model = "(1)"),
  cov_fit0$out |> mutate(model = "(2)")
) |>
  mutate(
    conf.low = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    N = 95
  ) |>
  ggregtab(ratio = .75) +
  facet_wrap(
    ~ paste0("log(",param,")"),
    labeller = label_parsed
  ) +
  scale_y_discrete(
    labels = c("Population (log)", "Polity (Avg.)", "Post-1950",
               "Constant")
  ) +
  labs(
    subtitle = str_wrap(
      "Table 2: Inverse Burr regression estimates with bootstrapped standard errors. Unadjusted fatality totals (1) compared with adjusted fatalities (2) as the outcome.",
      width = 80
    )
  )
saveit("_regtab2")
```


## Tabulations reported in-text

The share of all war deaths the top wars are responsible for.

```{r}
wars |>
  mutate(
    pct = fat / sum(fat)
  ) |>
  arrange(-pct) |>
  mutate(
    cumpct = cumsum(pct)
  )
```

